Types from data: \textnormal{Making structured data first-class citizens in F\#}

### 2016-02-09 pldi-2016/paper.tex
Final edits and fix bibtex

\documentclass[10pt,clearpagebib]{sigplanconf}

\usepackage{listings}
\usepackage{enumitem}
\usepackage[hyphens]{url}
\usepackage[svgnames]{xcolor}
\definecolor{lcl}{RGB}{140,0,100}
\usepackage[colorlinks=true,breaklinks,draft=false]{hyperref}
\hypersetup{urlcolor=lcl,linkcolor=lcl,citecolor=lcl}
\newcommand{\doi}[1]{doi:~\href{http://dx.doi.org/#1}{\Hurl{#1}}}
\usepackage{amssymb}
\usepackage{multicol}
\usepackage{xcolor}
\usepackage{graphicx}
\usepackage[fleqn]{amsmath}
\usepackage{graphics} 
\usepackage{stmaryrd}
\usepackage{amsthm}
\usepackage{ftnright}
\usepackage[T1]{fontenc}
\usepackage{semantic}
\usepackage{enumitem}
\usepackage[hang,flushmargin]{footmisc} 
\usepackage[notquote]{hanging}
\usepackage{flushend}

@@

\newtheorem{definition}{Definition}
\newtheorem{theorem}{Theorem}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{lemma}[theorem]{Lemma}

@@

One of the design motivations mentioned in Section~\ref{sec:providers} is that the inference
mechanism is predictable and stable. Informally this means that ...

In particular, we do 
not use probabilistic methods where adding an additional sample could completely change the 
shape of the type.

For the purpose of this section, assume conversion function \ident{int} and an error value
\ident{error} that propagates, i.e. $C[\ident{error}]\reduce\ident{error}$

\begin{theorem}[Stability of inference]

Given set of samples $d_1, \ldots, d_n$

And $\tau_1, e_1, L_1 = \sem{\semalt{d_1, \ldots, d_n}}$

And some code $L_1; x:\tau_1\vdash e : \tau$

~

Now, add $d_{n+1}$ and let
 $\tau_2, e_2, L_2 = \sem{\semalt{d_1, \ldots, d_n, d_{n+1}}}$

There is $e'$

such that $L_2; x:\tau_2\vdash e' : \tau$

and if for some $d$ it is the case that $e[x\leftarrow e_1~d] \reduce v$ then
also $e'[x\leftarrow e_2~d] \reduce v$

Obtained by changing

\begin{itemize}
\item[--]
$e$ to $\kvd{match}~e~\kvd{with}~\ident{Some}(v) \rightarrow v~|~\ident{None} \rightarrow \ident{wrong}$
\item[--]
$e$ to $e.\ident{Member}$ where $\ident{Member} = \tytagof(\sigma)$ for some $\sigma$
\item[--]
$e$ to $\ident{int}(e)$
\end{itemize}
\end{theorem}
\begin{proof}

Primitive type - can turn into float or into any with the primitive 
type exposed as one of the members (because it has unique tag), possibly turned to float

Record - introduce new fields, change types of fields - by induction
or not record, in which case it becomes member via any

List - element type change or becomes member of any

Any - ok

nullable - stays nullable, or if they're not compatible, part of any

\end{proof}

A full treatment of issue (2) would likely require more space than
available. However, we would like to add a brief section that makes the
following point: When a new sample is added to an existing set of samples,
the inferred shape can change only in certain limited ways (i.e. certain
parts may become optional). The fact that this change of shape is limited
shows that the inference algorithm is "stable" (users only need to do small
changes in their code). This is also relevant for the practical problem of
"choosing a representative sample". If an initial sample is not
representative and is refined later, the user will see which inferred types
became optional - and can adjust the program accordingly.

@@

We connect two lines of research: the work: inferring types for real-world data sources
and the integration of external data into statically-typed programming langauges.

Early work on F\# Data happened together with the development of the type providers mechanism in F\# 
\cite{fsharp-typeprov,fsharp-typeprov-ddfp}, but previous publications on type providers focused on 
the general mechanism \cite{idris-tp,liteq} rather than on concrete type providers. This paper 
is novel in that it discusses the theory and safety of a concrete type provider from the programming
language perspective.

\paragraph{Extending the type systems.} 
Previous work that integrates external data, such as XML \cite{xduce,xduce-ml} and databases 
\cite{links,linq}, into a programming language, required the user to define the schema
or had an ad-hoc extension that reads the schema.

C$\omega$ \cite{comega-xs} bears similarity to F\# Data. Like many other
languages, it extends a host language with advanced types similar to our
structural shapes (including similar heterogeneous collections), but it 
does not infer the types from a sample. In contrast, in F\# Data 
we have avoided extending F\# at all, and the simplicity of the Foo 
target language shows we have avoided placing strong requirements 
(such as structural records) on the host language. Instead we provide nominal 
types based on structure, rather than adding an advanced
system of structural types into the host language.

@@

\acks
We would like to thank to the F\# Data contributors on GitHub and other colleagues working 
on type providers, including Jomo Fisher, Keith Battocchi and Kenji Takeda.

>>>>

\documentclass[10pt]{sigplanconf}

\usepackage{listings}
\usepackage{enumitem}
\usepackage[hyphens]{url}
\usepackage[svgnames]{xcolor}
\definecolor{lcl}{RGB}{140,0,100}
\usepackage[colorlinks=true,breaklinks,draft=false]{hyperref}
\hypersetup{urlcolor=lcl,linkcolor=lcl,citecolor=lcl}
\newcommand{\doi}[1]{doi:~\href{http://dx.doi.org/#1}{\Hurl{#1}}}
\usepackage{amssymb}
\usepackage{multicol}
\usepackage{xcolor}
\usepackage{graphicx}
\usepackage[fleqn]{amsmath}
\usepackage{graphics} 
\usepackage{stmaryrd}
\usepackage{amsthm}
\usepackage{ftnright}
\usepackage[T1]{fontenc}
\usepackage{semantic}
\usepackage{enumitem}
\usepackage[hang,flushmargin]{footmisc} 
\usepackage[notquote]{hanging}
\usepackage{flushend}
\usepackage{enumerate}
\usepackage{etoolbox}

@@

\newtheorem{definition}{Definition}
\newtheorem{theorem}{Theorem}
\newtheorem{remark}{Remark}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{lemma}[theorem]{Lemma}
\newenvironment{proofoutline}
 {\renewcommand\qedsymbol{}\proof[Proof outline]}
 {\endproof}

@@

As discussed in \S\ref{sec:providers}, our inference algorithm is designed to be predictable
and stable. When a user writes a program using the provided type and then adds another sample 
(e.g.~with more missing values), they should not need to restructure their program. 
For this reason, we keep the algorithm simple. For example, we do not use probabilistic methods to 
asses similarity of record types, because a small change in the sample could cause a large change 
in the provided types.

We leave a general theory of stability and predictability of type providers to future work, but
we formalize a brief observation in this section. Say we write a program using a provided type
that is based on a collection of samples. When a new sample is added, the program can be modified
to run as before with only small local changes.

For the purpose of this section, assume that the Foo calculus also contains an \ident{exn} 
value representing a runtime exception that propagates in the usual way, 
i.e.~$C[\ident{exn}]\reduce\ident{exn}$, and also a conversion function \ident{int} that
turns floating-point number into an integer.

\begin{remark}[Stability of inference]
Assume we have a set of samples $d_1, \ldots, d_n$, provided type based on the samples
$\tau_1, e_1, L_1 = \sem{\semalt{d_1, \ldots, d_n}}$ and some user code $e$ written using
the provided type, such that $L_1; x:\tau_1\vdash e : \tau$.

Next, we add a new sample $d_{n+1}$ and consider a new provided type 
$\tau_2, e_2, L_2 = \sem{\semalt{d_1, \ldots, d_n, d_{n+1}}}$.

Now there exists $e'$ such that $L_2; x:\tau_2\vdash e' : \tau$ and if
for some $d$ it is the case that $e[x\leftarrow e_1~d] \reduce v$ then
also $e'[x\leftarrow e_2~d] \reduce v$.

Such $e'$ is obtained by applying one or more transformations to sub-expressions of $e$
that transform:
%
\begin{enumerate}[(i)]
\item
$C[e]$ to $C[\kvd{match}~e~\kvd{with}~\ident{Some}(v) \rightarrow v~|~\ident{None} \rightarrow \ident{exn}]$
\item
$C[e]$ to $C[e.\ident{M}]$ where $\ident{M} = \tytagof(\sigma)$ for some $\sigma$
\item
$C[e]$ to $C[\ident{int}(e)]$
\end{enumerate}
\end{remark}

\begin{proof}
For each case in the type provision (Figure~\ref{fig:tp-generation}) an original shape $\sigma$
may be replaced by a less preferred shape $\sigma'$. The user code can always be transformed
to use the newly provided shape:

\begin{itemize}
\item[--] Primitive shapes can become nullable (i), \ident{int} can become \ident{float} (iii)
  or can become a part of labelled top type (ii).
\item[--] Record shape fields can change shape (recursively) and record may become a part
  of labelled top type (ii).
\item[--] For list and nullable shape, the shape of the value may change (we apply the 
  transformations recursively).
\item[--] For any shape, the original code will continue to work (label is never removed).
\end{itemize}
\vspace{-1.5em}
\end{proof}
%
\noindent
Intuitively, the first transformation is needed when the new sample makes a type optional.
This happens when it contains a \kvd{null} value or a record that does not contain a field 
that all previous samples have. The second transformation is needed when a shape $\sigma$ 
becomes $\kvd{any}\langle\sigma, \ldots\rangle$ and the third one is needed when $\ident{int}$
becomes \ident{float}.

This property also underlines a common way of handling errors when using F\# Data type providers.
When a program fails on some input, the input can be added as another sample. This makes some
fields optional and the code can be updated accordingly, using a variation of (i) that uses
an appropriate default value rather than throwing an exception.

@@

The F\# Data library connects two lines of research that have been previously disconnected. The first is 
extending the type systems of programming languages to accommodate external data sources and the second
is inferring types for real-world data sources.

The type provider mechanism has been introduced in F\# \cite{fsharp-typeprov,fsharp-typeprov-ddfp}, 
added to Idris  \cite{idris-tp} and used in areas such as semantic web \cite{liteq}. The F\# Data 
library has been developed as part of the early F\# type provider research, but previous 
publications focused on the general mechanisms. This paper is novel in that it shows the 
programming language theory behind a concrete type providers.

\paragraph{Extending the type systems.} 
Several systems integrate external data into a programming language. Those include 
XML \cite{xduce,xduce-ml} and databases \cite{links}. In both of these, the system requires
the user to explicitly define the schema (using the host language) or it has an ad-hoc extension 
that reads the schema (\emph{e.g.}~from a database). LINQ \cite{linq} is more general, but relies
on code generation when importing the schema.

The work that is most similar to F\# Data is the data integration in C$\omega$ \cite{comega-xs}.
It extends C\# language with types similar to our structural types 
(including nullable types, choices with subtyping and heterogeneous collections with multiplicities).
However, C$\omega$ does not infer the types from samples and extends the type system of the host
language (rather than using a general purpose embedding mechanism).

In contrast, in F\# Data 
we have avoided extending F\# at all, and the simplicity of the Foo 
target language shows we have avoided placing strong requirements 
(such as structural records) on the host language. Instead we provide nominal 
types based on structure, rather than adding an advanced
system of structural types into the host language.

@@

\acks
We would like to thank to the F\# Data contributors on GitHub and other colleagues working 
on type providers, including Jomo Fisher, Keith Battocchi and Kenji Takeda. We are also
grateful to anonymous reviewers of the paper for their valuable feedback.
